A density ratio framework for evaluating the utility of synthetic data

### 2024-04-26 paper/paper.tex
Update paper

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A density ratio framework for evaluating the utility of synthetic data},
  pdfauthor={Thom Benjamin Volker; Peter-Paul de Wolf; Erik-Jan van Kesteren},
  pdfkeywords={TODO, TODO},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{setspace}
\doublespacing
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Density ratios for utility }
\title{A density ratio framework for evaluating the utility of synthetic
data}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Thom Benjamin
Volker}~\orcidlink{0000-0002-2408-7820}\\Methodology and Statistics
\textbar{} Methodology\\Utrecht University \textbar{} Statistics
Netherlands\\Utrecht,\ 3584CH\\\href{mailto:t.b.volker@uu.nl}{t.b.volker@uu.nl}\asep\textbf{Peter-Paul
de Wolf}\\Methodology\\Statistics Netherlands\\The
Hague,\ 2490HA\\\href{mailto:pp.dewolf@cbs.nl}{pp.dewolf@cbs.nl}\asep\textbf{Erik-Jan
van Kesteren}\\Methodology and Statistics\\Utrecht
University\\Utrecht,\ 3584CH\\\href{mailto:e.vankesteren1@uu.nl}{e.vankesteren1@uu.nl}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
TODO \sep 
TODO

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

Openly accessible research data accelerates scientific progress
tremendously {[}ADD REFERENCES{]}. Open data allows third-party
researchers to answer research questions with already collected data,
substantially lowering the costs of research. Sharing data in
combination with code allows others to validate research findings and
build upon the work. Students can benefit from open data, as it fosters
education with realistic data, as well as the general public, through
stimulating citizen science projects. However, openly available research
data is often at odds with privacy constraints on the dissemination of
the data. The information may cause harm to individuals or organizations
if made publicly available. Additionally, sharing their information
openly may withhold them from participating in future research. These
constraints have been named among the biggest hurdles in the advancement
of computational social science
(\protect\hyperlink{ref-lazer_css_2009}{Lazer et al. 2009}), and among
top reasons for companies to not share their data with researchers
(\protect\hyperlink{ref-fpf_2017}{Future of Privacy Forum 2017}). To
overcome these obstacles, data providers can employ a suite of different
disclosure limitation techniques before sharing data, for example
top-coding, record-swapping or adding noise (e.g.,
\protect\hyperlink{ref-hundepool_disclosure_2012}{Hundepool et al.
2012}; \protect\hyperlink{ref-willenborg_elements_2012}{Willenborg and
De Waal 2012}).

Recently, synthetic data has gained substantial traction as a means to
disclosure limitation. National statistical institutes and other
government agencies have adopted the synthetic data framework to
facilitate the use of their data (e.g.,
\protect\hyperlink{ref-SIPP_Beta_2006}{Abowd, Stinson, and Benedetto
2006}; \protect\hyperlink{ref-hawala_synthetic_2008}{Hawala 2008};
\protect\hyperlink{ref-drechsler2012}{Drechsler 2012}). Researchers
started using synthetic data to train machine learning models
(\protect\hyperlink{ref-nikolenko2021}{Nikolenko 2021}), and began to
share synthetic versions of their research data to comply with open
science standards (e.g., \protect\hyperlink{ref-vandewiel2023}{van de
Wiel et al. 2023}; \protect\hyperlink{ref-obermeyer2019}{Obermeyer et
al. 2019}; \protect\hyperlink{ref-zettler2021}{Zettler et al. 2021}).
The idea of synthetic data is to replace some, or all, of the observed
values in a data set by synthetic records that are generated from some
model (e.g., \protect\hyperlink{ref-little_statistical_1993}{Little
1993}; \protect\hyperlink{ref-rubin_statistical_1993}{Rubin 1993};
\protect\hyperlink{ref-drechsler2011synthetic}{Drechsler 2011}). If only
some values are replaced, disclosure risks can be reduced because the
sensitive or identifying values do not correspond to their true values
anymore. If all values are replaced, there is also no one-to-one mapping
between the original and the synthetic data, further reducing the
disclosure risk.

Many different methods have been proposed to generate synthetic data.
Traditionally, these were closely connected to methods used for multiple
imputation of missing data, such as joint modelling {[}misdc2003{]},
sequential regressions (\protect\hyperlink{ref-nowok2016}{Nowok, Raab,
and Dibben 2016}) or fully conditional specification
(\protect\hyperlink{ref-drechsler2011empirical}{Drechsler and Reiter
2011}; \protect\hyperlink{ref-volker2021}{Volker and Vink 2021}). The
flexibility of sequential regressions and fully conditional
specification is commonly combined with non-parametric imputation
models, such as classification and regression trees
(\protect\hyperlink{ref-reiter2005}{Reiter 2005}), random forests
(\protect\hyperlink{ref-caiola2010}{Caiola and Reiter 2010}) or support
vector machines (\protect\hyperlink{ref-drechsler2010}{Drechsler 2010}),
to avoid distributional assumptions and easily model non-linear
relationships. Recently, significant improvements in generative
modelling sparked the scientific interest in synthetic data in the
computer science community, leading to novel synthesis methods (e.g.,
\protect\hyperlink{ref-patki2016}{Patki, Wedge, and Veeramachaneni
2016}; \protect\hyperlink{ref-xu_ctgan_2019}{Xu et al. 2019}). Combined
with work on formal privacy guarantees such as differential privacy,
this resulted in new models that explicitly control the level of privacy
risk in synthesis methods
(\protect\hyperlink{ref-jordon2018pategan}{Jordon, Yoon, and Schaar
2019}; \protect\hyperlink{ref-Torkzadehmahani2019}{Torkzadehmahani,
Kairouz, and Paten 2019}).

\textbf{\emph{TODO}}

All methods for statistical disclosure limitation alter the data before
these are provided to the public. By doing so, the utility of the
provided data is always lower than the utility of the original data,
because some of the information in the data is sacrificed to protect the
privacy of the respondents. The questions that naturally arise are how
much information in the original data is actually sacrificed, and how
useful the provided data are? Answering this question allows researchers
to decide what the altered data can and cannot be used for, and to
evaluate the worth of conclusions drawn on the basis of these data.
After all, inferences from the altered data are valid only up to the
extent that the perturbation methods approximate the true
data-generating mechanism. For data providers, a detailed assessment of
the quality of the altered data can guide the procedure of altering the
data. Statistical disclosure limitation is often an iterative process:
some disclosure limitation technique is applied on the data, after which
the result is investigated and modifications are made to applied
process. Good measures of data quality are essential to determine the
appropriate mechanisms used to protect the data, and can help to improve
the utility of the data that will be disseminated.

In the statistical disclosure control literature, two different branches
of utility measures have been distinguished: specific utility measures
and general utility measures. \emph{Add one/two sentences on the merit
of visualization when assessing utility of altered data.} Specific
utility measures focus on similarity of results obtained from analyses
performed on the altered data and the original data. For example, after
fitting the same analysis model on both data sets, one can calculate the
confidence interval overlap of the estimated parameters
(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}).
Alternative measures are ellipsoidal overlap
(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}), which
extends to confidence interval overlap to a measure that addresses the
joint distribution of all model parameters simultaneous, the
standardized absolute difference between estimates
(\protect\hyperlink{ref-snoke_utility_2018}{Snoke et al. 2018}), and the
ratio of estimates for tabular count data
(\protect\hyperlink{ref-taub2020impact}{\textbf{taub2020impact?}}). As
these measures quantify similarity between estimates from analyses
performed on the observed and altered data, they are informed only to
the extent that data users will recreate those analyses. This can be
highly useful if the data is provided for reproducibility purposes
(e.g., for third parties to evaluate analysis scripts). However, the
goal of distributing the protected data is often to allow researchers to
do novel research with the data. In many practical situations, data
providers thus have have only limited knowledge on the analyses that
will be performed with the altered data. Covering the entire set of
potentially relevant analyses is therefore not feasible. If it was, the
data providers could simply report the (potentially privacy-protected)
results of those analyses performed on the real data, so that access to
the (perturbed) data no longer yields additional benefits (for a similar
argument, see \protect\hyperlink{ref-drechsler_utility_2022}{Drechsler
2022}). Additionally, similarity between results on the analyses that
have been performed gives no guarantee that the results will also be
similar for other analyses. Hence, when determining how useful the
altered data is for novel research, specific utility measures are only
of limited use.

General utility measures attempt to capture how similar the multivariate
distributions of the observed and altered data are. This can be done by,
for instance, estimating the Kullback-Leibler divergence between the
distributions of the observed and altered data
(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}). An
alternative strategy is to try to discriminate between the observed and
altered data, as is done with the \(pMSE\)
(\protect\hyperlink{ref-snoke_utility_2018}{Snoke et al. 2018};
\protect\hyperlink{ref-woo_utility_2009}{\textbf{woo\_utility\_2009?}}).
In essence, the \(pMSE\) quantifies how well one can predict whether
observations are from the observed or the altered data. If better one
can do this, the more pronounced the differences between the observed
and altered data ought to be. However, various authors have criticized
general utility measures for being too broad. That is, important
discrepancies between the real and altered data might be missed, and an
altered data set that is good in general (i.e., has high global utility)
might still provide results that are far from the truth for some
analyses (see, e.g.,
\protect\hyperlink{ref-drechsler_utility_2022}{Drechsler 2022}).
Additionally, it is not straightforward to determine which prediction
model to use for calculating the \(pMSE\). Specifying a good prediction
model in itself may be a challenging task, especially when the number of
variables is large. When good models are available, different models, or
even different choices of hyperparameters, may yield different results,
potentially rendering ambiguity with respect to which altered data set
is best. Lastly, the output of global utility measures can be hard to
interpret, and say little about the regions in which the synthetic data
do not resemble the true data accurately enough. That is, they give
little guidance on how the quality of the altered data can be improved.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\textbf{Section 6: our contribution}

\emph{Moet nog verder uitgewerkt worden}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We introduce density ratio estimation to the field of statistical
  disclosure control. Short remark on the idea that density ratio
  estimation is a complicated endeavor, especially if the goal is to
  compare distinct densities. Having to estimate just a single density
  (ratio) is generally much easier.
\item
  Note that density ratio estimation can capture specific and general
  utility measures into a common framework by being applicable on the
  level of the entire data, but also on the subset of variables that is
  relevant in an analysis. Additionally, note that confidence interval
  overlap, ellipsoidal overlap, but also \(pMSE\) and Kullback-Leibler
  divergence, are closely related to density ratio estimation, and can
  be considered from this perspective.
\item
  Create a new utility metric based on density ratio estimation
  (probability with respect to some reference distribution as in
  permutation testing).
\item
  Because density ratio estimation can be difficult when there are many
  variables, we use dimension reduction techniques to capture most of
  the variability in the data in fewer dimensions on which density ratio
  estimation can be applied. A by-product of this is that the
  lower-dimensional subspace allows to create visualizations of
  deviations from the observed data.
\item
  Perform a simulation study to give indications about which methods to
  use (think about how to do this).
\item
  Implement all this in an R-package
\end{enumerate}

\textbf{Section 6: outline article}

In the next section, we describe density ratio estimation and discuss
how this method can be used as to measure utility. Subsequently, we
provide multiple examples that show how density ratio estimation works
in the context of evaluating the quality of synthetic data. Hereafter,
we show in multiple simulations that the method is superior (HOPEFULLY)
to current global utility measures as the \(pMSE\). Lastly, we discuss
the advantages and disadvantages of density ratio estimation as a
utility measure.

\hypertarget{methodology}{%
\section{Methodology}\label{methodology}}

@Peter-Paul: Eventueel een korte beschrijving van data perturbation
techniques/synthetic data generation hier. Denk je dat dit wat toevoegt
hier?

\textbf{Section 1: density ratio estimation}

In essence, the goal of utility measures is to quantify the similarity
between the multivariate distribution of the observed data with the
distribution of the altered data. If the used data perturbation
techniques, or synthetic data generation models, approximate the
distribution of the real data sufficiently, these distributions should
be highly similar, and analyses on the two data sets should give similar
results. However, estimating the probability distribution of a data set
is known to be one of the most complicated challenges in statistics
{[}E.G. Vapnik 1998{]}. Estimating the probability distribution for both
observed and altered data can lead to errors in both, artificially
magnifying discrepancies between the two. Hence, subsequent comparisons
will be affected by these errors. The procedure can be simplified by
using density ratio estimation, because this only requires to estimate a
single density.

Introduce density ratio estimation as a utility measure. What does this
measure mean/how to interpret it. How to make decisions based on this
measure.

Say something on whether (and if so, how) categorical variables can be
incorporated as well.

\textbf{Section 3: theoretical comparison with conventional approaches
for general utility assessment}

Relate density ratio estimation to specific and general utility
measures. Pick one/two specific utility measures and relate these to
density ratio estimation (ratio of estimates seems straightforward, as
well as confidence interval overlap and ellipsoidal overlap).

Relate density ratio estimation to \(pMSE\) and KL divergence (to some
extent, both are generalizations of density ratio estimation, or at
least are conceptually similar). Give some more information on the
\(pMSE\), describe what it shortcomings are. The quality of the \(pMSE\)
highly depends on the model used to calculate the propensity scores.
Perhaps give an example of logistic regression, which basically
estimates whether the conditional mean of the observed and altered data
is the same. Explain how density ratio estimation can overcome the
shortcomings of the previously mentioned methods.

\textbf{Section 4: Dimension reduction for utility}

The difficulty of density ratio estimation increases with the
dimensionality of the data. Therefore, we follow previous
recommendations to incorporate dimensionality reduction techniques in
density ratio estimation.

>>>>

\newcommand{\Xobs}{\mathbf{X}_{\text{obs}}}
\newcommand{\Xsyn}{\mathbf{X}_{\text{syn}}}
\newcommand{\dist}{\mathcal{P}}
\newcommand{\pobs}{p_{\text{obs}}}
\newcommand{\psyn}{p_{\text{syn}}}
\newcommand{\nobs}{n_{\text{obs}}}
\newcommand{\nsyn}{n_{\text{syn}}}
\newcommand{\bx}{\mathbf{x}}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A density ratio framework for evaluating the utility of synthetic data},
  pdfauthor={Thom Benjamin Volker; Peter-Paul de Wolf; Erik-Jan van Kesteren},
  pdfkeywords={synthetic data, utility, density ratio, privacy,
disclosure limitation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{setspace}
\doublespacing
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Density ratios for utility }
\title{A density ratio framework for evaluating the utility of synthetic
data}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Thom Benjamin
Volker}~\orcidlink{0000-0002-2408-7820}\\Methodology and Statistics
\textbar{} Methodology\\Utrecht University \textbar{} Statistics
Netherlands\\Utrecht,\ 3584CH\\\href{mailto:t.b.volker@uu.nl}{t.b.volker@uu.nl}\asep\textbf{Peter-Paul
de Wolf}\\Methodology\\Statistics Netherlands\\The
Hague,\ 2490HA\\\href{mailto:pp.dewolf@cbs.nl}{pp.dewolf@cbs.nl}\asep\textbf{Erik-Jan
van Kesteren}\\Methodology and Statistics\\Utrecht
University\\Utrecht,\ 3584CH\\\href{mailto:e.vankesteren1@uu.nl}{e.vankesteren1@uu.nl}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
synthetic data, utility, density ratio, privacy, disclosure
limitation \sep 
synthetic data, utility, density ratio, privacy, disclosure limitation

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

Openly accessible research data accelerates scientific progress
tremendously. Open data allows third-party researchers to answer
research questions with already collected data, freeing up resources
that would otherwise be devoted to data collection
(\protect\hyperlink{ref-ramachandran_open_2021}{Ramachandran, Bugbee,
and Murphy 2021}). Sharing data in combination with code allows others
to validate research findings and build upon the work
(\protect\hyperlink{ref-obels_analysis_2020}{Obels et al. 2020};
\protect\hyperlink{ref-crosas_automating_2015}{Crosas et al. 2015}).
Students can benefit from open data, as it fosters education with
realistic data (\protect\hyperlink{ref-atenas_open_2015}{Atenas,
Havemann, and Priego 2015}), as well as the general public, through
stimulating citizen science projects
(\protect\hyperlink{ref-newman_future_2012}{Newman et al. 2012}).
However, making data openly available is often (rightfully) hampered by
official legislation, like the General Data Protection Regulation (GDPR;
\protect\hyperlink{ref-gdpr}{European Parliament and Council of the
European Union 2016}), and general privacy concerns. In the worst case,
sharing data may cause harm to individuals or organizations, which may
withhold these entities from participating in future research. These
privacy constraints have been named among the biggest hurdles in the
advancement of computational social science
(\protect\hyperlink{ref-lazer_css_2009}{Lazer et al. 2009}), and among
top reasons for companies to not share their data with researchers
(\protect\hyperlink{ref-fpf_2017}{Future of Privacy Forum 2017}).

Multiple approaches exist to balance the benefits of open data with
potential privacy risks. Traditionally, data providers employed a suite
of different disclosure limitation techniques before sharing the data,
such as top-coding, record-swapping or adding noise (e.g.,
\protect\hyperlink{ref-hundepool_disclosure_2012}{Hundepool et al.
2012}; \protect\hyperlink{ref-willenborg_elements_2001}{Willenborg and
De Waal 2001}). More recently, synthetic data has gained traction as a
means to disseminate private data
(\protect\hyperlink{ref-SIPP_Beta_2006}{Abowd, Stinson, and Benedetto
2006}; \protect\hyperlink{ref-hawala_synthetic_2008}{Hawala 2008};
\protect\hyperlink{ref-drechsler2012}{Drechsler 2012};
\protect\hyperlink{ref-vandewiel2023}{van de Wiel et al. 2023};
\protect\hyperlink{ref-obermeyer2019}{Obermeyer et al. 2019};
\protect\hyperlink{ref-zettler2021}{Zettler et al. 2021}), although the
conceptual framework traces back to the previous century
(\protect\hyperlink{ref-little_statistical_1993}{Little 1993};
\protect\hyperlink{ref-rubin_statistical_1993}{Rubin 1993}). Simply put,
the idea of synthetic data is to replace some, or all, of the observed
values in a data set by synthetic values that are generated from some
model (e.g., \protect\hyperlink{ref-drechsler2011synthetic}{Drechsler
2011}). If only some values are replaced, disclosure risks can be
reduced because the sensitive or identifying values do not correspond to
their true values anymore. If all values are replaced, there is also no
one-to-one mapping between the original and the synthetic data, further
reducing the disclosure risk. However, an increase in privacy typically
comes at the cost of a decrease in utility. As more of the data is
altered, the quality of the released data becomes more sensitive to the
suitability of the generative model. Regardless of the approach to
disclosure limitation, if the technique used to generate or alter the
data does not align with the intricacies of the problem at hand, the
utility of the released data will be further reduced than necessary.

Given that all disclosure limitation techniques reduce the utility of
the data, the challenge that arises is how to determine whether the
released data still has acceptable utility. Alternatively, one might
consider different disclosure limitation techniques that all satisfy the
defined privacy restrictions, and employ the one which yields data with
the highest utility. That is, given a set of methods that all meet the
privacy restrictions, one may aim to maximize the utility of the
released data. Both strategies require a reliable and encompassing
measure of data utility that allows to evaluate the quality of the
released data, and that allows to compare different disclosure
limitation techniques and/or synthesis models in terms of utility.
Moreover, adequate utility measures often guide the synthesis process,
by providing detailed feedback on important discrepancies between the
original and synthetic data. Lastly, good utility measures help the data
user in determining what the synthetic data can and cannot be used for.

In the synthetic data field, three classes of utility measures have been
distinguished (see \protect\hyperlink{ref-drechsler2023}{Drechsler and
Haensch 2023} for a thorough review): fit-for-purpose measures,
analysis-specific utility measures and global utility measures.
Fit-for-purpose measures are often the first step in assessing the
quality of the synthetic data. These typically involve comparing the
univariate distributions of the observed and synthetic data (for example
using visualization techniques or goodness-of-fit measures). Although
these measures provide an initial impression of the quality of the
synthesis models used, this picture is by definition limited, because
only one or two variables are assessed at the same time. Hence, complex
relationships between variables will always be out of scope. Such
relationships may be captured by analysis-specific utility measures,
which quantify whether analyses on synthetic data provide results that
are comparable to results from the same analysis performed on the
observed data. These measures can, for example, evaluate how similar the
coefficients of a regression model are (e.g., using the confidence
interval overlap; \protect\hyperlink{ref-karr_utility_2006}{Karr et al.
2006}), or whether prediction models trained on the synthetic and
observed data perform comparably in terms of evaluation metrics.
However, analysis-specific utility generally does not carry over: high
specific utility for one analysis does not at all imply high utility for
another analysis. Since data providers typically do not know which
analyses will be performed with the synthetic data, it is impossible to
provide analysis-specific utility measures for all potentially relevant
analyses (see also
\protect\hyperlink{ref-drechsler_utility_2022}{Drechsler 2022}).

Global utility measures may overcome the shortcomings of the previous
approaches, as they evaluate the discrepancy between the entire
multivariate distribution of the observed and synthetic data. As such,
global utility measures yields the most promising class of utility
measures, because if the observed and synthetic data have similar
(multivariate) distributions, all potential analyses should yield
similar results. Global utility can be evaluated using some divergence
measure, such as the Kullback-Leibler divergence
(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}), or by
evaluating whether the observed and synthetic data are distinguishable
using a classification model (a technique called \(pMSE\);
\protect\hyperlink{ref-Woo_global_2009}{Woo et al. 2009};
\protect\hyperlink{ref-snoke_utility_2018}{Snoke et al. 2018}). However,
a common critique of global utility measures is that they tend to be too
general (\protect\hyperlink{ref-drechsler_utility_2022}{Drechsler
2022}). That is, analyses on a synthetic data set that is overall quite
similar to the observed data (i.e., has high global utility), may still
yield results that are far from the results obtained from the real data.
Also, commonly used methods for estimating the \(pMSE\), as logistic
regression and classification and regression trees, tend to become less
reliable as the dimensionality of the data increases, and are vulnerable
to model misspecification
(\protect\hyperlink{ref-drechsler_utility_2022}{Drechsler 2022}).
Lastly, the output of global utility measures can be hard to interpret,
and say little about the regions in which the synthetic data do not
resemble the true data accurately enough.

To overcome the issues related to traditional global utility measures,
we propose to use the density ratio estimation framework
(\protect\hyperlink{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki,
and Kanamori 2012a}) as a way of evaluating utility. Intuitively, if two
data sets have similar multivariate distributions, the density ratio
should be close to one over the range of the data. If the distributions
of the observed and synthetic data are very different, the density ratio
should be far from one at those regions where the distributions differ.
As density estimation is known to be a difficult problem, the density
ratio estimation framework provides techniques to directly estimate the
density ratio, rather than the two separate densities, in a
non-parametric way
(\protect\hyperlink{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki,
and Kanamori 2012a}). These non-parametric estimation techniques come
with automatic model specification, which mitigates the issue of model
specification. This functionality is implemented in the
\texttt{R}-package \texttt{densityratio}
(\protect\hyperlink{ref-densityratio}{Volker 2023}). Importantly, the
density ratio is estimated over the entire range of the data, which
provides measures of utility at every (possible) point in the data
space. This point-specific quantification of utility turns out to be a
useful side-product, as it allows to reweigh analyses on synthetic data
when further improving the utility directly is not possible.

In the remainder of the article, we introduce the density ratio
framework and the associated estimation techniques, and connect the
framework to traditional utility measures as the \(pMSE\) and the
Kullback-Leibler divergence. We then present simulations to demonstrate
the performance of density ratio estimation in stylized settings and
compare it to traditional utility measures. Subsequently, we apply
density ratio estimation in a case study where we evaluate the utility
of multiple synthetic versions of the U.S. Current Population Survey. We
conclude with a discussion of the results, highlight the strengths and
weaknesses of the density ratio framework, and provide recommendations
for future research.

\hypertarget{background}{%
\section{Background}\label{background}}

Over the years, many methods have been introduced to generate synthetic
data, all with the aim of providing a suitable balance between privacy
and utility. These methods can be relatively simple, such as a sequence
of generalized linear models (e.g.,
\protect\hyperlink{ref-reiter_releasing_2004}{Reiter 2004}), or as
complex as deep learning models with thousands of parameters (e.g.,
\protect\hyperlink{ref-xu_ctgan_2019}{Xu et al. 2019}), with many
options in between. However, the complexity of the generation process is
not necessarily a good indicator of the quality of the synthetic data.
That is, relatively simple methods could still capture the most
important aspects of the data that complex methods fail to capture (and
vice versa). Hence, data providers typically do not know which synthesis
method will provide the highest utility a priori, and might compare
multiple synthesis strategies to determine which data set will be
released. Good global utility measures can help in this process, by
allowing to quantify the quality of the candidate synthetic data
sets.\footnote{We focus on global utility measures, because in many
  situations the data provider does not know which analysis will be
  performed with the synthetic data. If the data provider knows for
  which purposes the data will be used, analysis-specific utility
  measures may be more informative.} Moreover, such global utility
measures may guide the synthesis process itself, if they provide
sufficiently specific information about the degree of misfit of the
synthetic data. In the upcoming section, we provide an overview of
existing global utility measures, and introduce the density ratio
framework as an encompassing approach to evaluating global utility.

\hypertarget{existing-global-utility-measures}{%
\subsection{Existing global utility
measures}\label{existing-global-utility-measures}}

Global utility measures typically attempt to quantify the distributional
similarity between the observed and synthetic data samples. The
intuition is that if two data sets have similar distributions, the data
sets can be used for the same purposes, and analyses on the two data
sets should yield similar results. A common way to formalize
distributional similarity is through the Kullback-Leibler (KL)
divergence, as proposed in Karr et al.
(\protect\hyperlink{ref-karr_utility_2006}{2006}). The KL-divergence
measures the relative entropy from the probability distribution of the
observed data \(\pobs(\bx)\) to the probability distribution of the
synthetic data \(\psyn(\bx)\) (with
\(\bx \in \mathbb{R}^{n \times d}\)), and is defined as
\begin{equation}\protect\hypertarget{eq-kl-div}{}{
D_{\text{KL}}(\psyn || \pobs) = 
\int \psyn(\bx) \log \frac{\psyn(\bx)}{\pobs(\bx)} \text{d} \bx.
}\label{eq-kl-div}\end{equation} Karr et al.
(\protect\hyperlink{ref-karr_utility_2006}{2006}) argue that the
KL-divergence can be constructed from density estimators
\(\hat{p}_\text{obs}\) and \(\hat{p}_\text{syn}\), after which the
integral can be approximated using numerical quadrature. Alternatively,
if the observed and synthetic data are both (multivariate) normally
distributed, the KL-divergence can be computed analytically. Both
implementations might be unfeasible: the combination of density
estimation with numerical quadrature is cumbersome in high dimensional
settings, and assuming multivariate normality might be too restrictive.
Yet, we show in later sections that the density ratio estimation
framework gives rives to an alternative way of estimating the
KL-divergence that overcomes these challenges.
